---
title: "phylogenetic distance-primate examples"
output: html_document
date: "2026-05-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Closing the backdoor path by using phylogenetic distance to reconstruct variation between predictor and response variable 

In other words, partition the evolutionary relationship from the relationship between predictors (e.g., life history traits) and response variables (e.g., compositional patterns) we are interested in? 

We are going to plot the consensus phylogeny for 301 primate species and then use the tree to to model our unobserved confounds and close the backdoor path between our predictor and response variable. 

```{r consensus phylogeny for 301 primate species}
library(rethinking)
data(Primates301)
data(Primates301_nex)

# plot using ape package -- install.packages('ape') if needed
library(ape)
plot(ladderize(Primates301_nex), type="fan", font=1, no.margin=TRUE, label.offset=1, cex=0.5)
```

## Fitting the ordinary regression model to the primate data

```{r remove species with no group size, brain size, and body size data, echo=FALSE}
d <- Primates301
d$name <- as.character(d$name)
dstan <- d[complete.cases(d$group_size, d$body, d$brain),]
spp_obs <- dstan$name
```

## Make a list of standardised logged variables and pass it all to ulam
```{r model 14.9}
dat_list <- list(
  N_spp = nrow(dstan), #placeholder according to the number of species
  M = standardize(log(dstan$body)), #body mass
  B = standardize(log(dstan$brain)), #brain size
  G = standardize(log(dstan$group_size)), #group size
  Imat = diag(nrow(dstan)) #identity matrix
)

m14.9 <- ulam(
  alist(
    B ~ multi_normal(mu, SIGMA), #vector of species brain sizes
    mu <- a + bM*M + bG*G, #model equation 
    matrix[N_spp,N_spp]: SIGMA <- Imat * sigma_sq, #covariance matrix
    a ~ normal(0,1), #prior distribution for intercept
    c(bM, bG) ~ normal(0,0.5), #prior distribution for coefficient of body mass and group size
    sigma_sq ~ exponential(1) #prior distribution of variance 
  ), data=dat_list, chains=4, cores=4)
```
```{r check the model outputs}
precis(m14.9)
```

```{r compute the distance matrix for Brownian motion}
library(ape)
tree_trimmed <- keep.tip(Primates301_nex, spp_obs)
Rbm <- corBrownian(phy=tree_trimmed)
V <- vcv(Rbm)
Dmat <- cophenetic(tree_trimmed)
plot(Dmat, V, xlab="phylogenetic distance", ylab="covariance")
```
```{r plot images from the first kind of regression - Brownian motion}
image(V)
image(Dmat)
```
```{r convert the model to correlation matrix to estimate residual variance}

# put species in the right order
dat_list$V <- V[ spp_obs, spp_obs ]

# convert to correlation matrix
dat_list$R <- dat_list$V/max(V)

# Brownian motion model 
m14.10 <- ulam(
  alist(
    B ~ multi_normal(mu, SIGMA),
    mu <- a + bM*M + bG*G,
    matrix[N_spp, N_spp]: SIGMA <- R * sigma_sq,
           a ~ normal(0, 1),
           c(bM, bG) ~ normal(0, 0.5),
           sigma_sq ~ exponential (1)
    ), data=dat_list, chains=4, cores=4)

```

```{r plot the model outputs}
precis(m14.10)
```
```{r second phylogenetic regression - ornstein-uhlenbeck process}

# add scaled and reordered distance matrix
dat_list$Dmat <- Dmat[ spp_obs, spp_obs ] / max(Dmat)


m14.11 <- ulam(
  alist(
    B ~ multi_normal(mu, SIGMA),
    mu <- a + bM*M + bG*G,
    matrix[N_spp, N_spp]: SIGMA <- cov_GPL1(Dmat, etasq, rhosq, 0.01),
           a ~ normal(0, 1),
           c(bM, bG) ~ normal(0, 0.5),
           etasq ~ half_normal(1, 0.25),
           rhosq ~ half_normal(3, 0.25)
    ), data=dat_list, chains=4, cores=4)
```

```{r look at the model outputs}
precis(m14.11)
```

```{r}


```

